---
title: "Gene Set Testing"
author: "Claire Levy"
date: "March 11, 2016"
output: github_document
---


## Gene set tests of explant data

From the description of the `camera` function from limma:

"Test whether a set of genes is highly ranked relative to other genes in terms of differential expression, accounting for inter-gene correlation."

The user provides:

* a filtered set of expression data (the same set that I used for doing differential expression analysis)

* a design matrix (also the same as for DE analysis)

* a contrast, or, the parts of the expression data we want to compare (in our case V186 24 hours vs Mock 24hrs)

* a gene set of interest

## How CAMERA works

`camera` looks at how many of the genes that were expressed in the contrast we specified are also in the gene set we provided. The test result gives 

* The number of genes in the input data that were in the gene set

* a p-value that tells us whether the genes that are in the set are more differentially expressed compared with the complement ( i.e. our genes that _weren't_ in the set)

* The correlation between the genes that were in the set

* the direction of change, relative to the gene set. For example, if the direction is DOWN, genes that were in the set have lower DE than genes not in the set.

* if we test >1 gene set, we are given a corrected p-value (FDR)

## Test 1

For the first test I used the following gene set:

Probes from the epithelial cell microarray that I found to be DE in the V186 infected, 24 hr timepoint(V186.24) condition.

Here is the test result

```{r,echo=FALSE,message=FALSE, warning=FALSE}


####  SEE CAMERA_gene_set_testing_README for more on this process #####

library(limma)
library(lumi)
library(pander)
library(dplyr)


##Expression set to use

load("RdataOutput/expressedProbes.lumi.Rdata")

## design matrix is the same as for the explant analysis
load("RdataOutput/design.Rdata")

y<-exprs(expressedProbes.lumi)

#Using the DE probes from the epithelial cell MA as the set to compare against.


load("RdataOutput/CELLStt3.Rdata")#top table for cells V186.24 DE probes

#split CELLStt3 into "UP" and "DOWN" regulated probes

CELLStt3UP<-CELLStt3 %>% 
  filter(logFC>0)

CELLStt3DOWN<-CELLStt3 %>% 
  filter(logFC<0)



#set up the gene sets for index. There are 11076 probes (8873 unique ones) BUT 


gene.sets<-list("UP"=CELLStt3UP$ENTREZ_GENE_ID, "DOWN"=CELLStt3DOWN$ENTREZ_GENE_ID)



#IDs from eset: ENTREZ_IDs from the data I'm analyzing (not the pre-determined gene set)

identifiers = as.character(fData(expressedProbes.lumi)[,"ENTREZ_GENE_ID"])




#make index with gene.sets entreZs and eset IDs
#"list of integer vectors, each vector containing the #indices of a gene set in the vector identifiers."


index<-ids2indices(gene.sets = gene.sets, identifiers = identifiers)

#define a matrix of the contrast that you want the results for
contr<-makeContrasts(TreatV186.24-TreatMock.24, levels = design)

#put arguments into camera
cameraResult<-camera(y = y, index = index, design = design, contrast = contr )

pander(cameraResult)

```

This is weird.

## Test 2

Also DE probes from the CELLS experiment, V186 treatment, but they are for all 3 time points (3hr, 8hr, 24hr) and are separated by whether the direction of DE was up or down.


```{r,echo=FALSE,message=FALSE, warning=FALSE}

###### Repeat test but for gene sets, use DE probes from all v186 CELLS data, separated by direction of change and time point.

#Prepare the gene sets by getting the top tables for CELLSV186 UP and DOWN reg DE probes from each timepoint all into a single list.

#This is a list of the top tables for the 3 time points (in order 3, 8 and 24)from the CELLS experiment analysis.

load("RdataOutput/CELLSttList.Rdata")

#add names to the dfs in the list so I don't get confused
names(CELLSttList) <-c("CELLS_V186.3","CELLS_V186.8","CELLS_V186.24")

#functions to separate out the the up and down regulated probes, based on the sign of the data in the logFC column.

#at the same time, select just the column I'm interested in.
getUP<-function(df){
  df%>%
    filter(logFC>0)
}


getDOWN<-function(df){
  df%>%
    filter(logFC<0)
}


#get upreg
CELLSV186UP<-lapply(CELLSttList,FUN=getUP)

#get downreg      
CELLSV186DOWN<-lapply(CELLSttList,FUN=getDOWN)

##add directional annotation to CELLS list. These are lists of DE probes for the 3 time points, V186 only.

names(CELLSV186UP)<-paste(names(CELLSV186UP), "UP", sep= "")

CELLSV186UP<-bind_rows(CELLSV186UP)

names(CELLSV186DOWN)<-paste(names(CELLSV186DOWN), "DOWN", sep= "")

CELLSV186DOWN<-bind_rows(CELLSV186DOWN)

#make it into one list
CELLSV186DE<-list("UP" = CELLSV186UP$ENTREZ_GENE_ID,"DOWN" = CELLSV186DOWN$ENTREZ_GENE_ID)


```



```{r,echo=FALSE,message=FALSE, warning=FALSE}

##Expression set to use

y<-exprs(expressedProbes.lumi)

#IDs from eset
identifiers = as.character(fData(expressedProbes.lumi)[,"ENTREZ_GENE_ID"])



#I am using my list of DE probes from the 3 different timepoints for CELLSV186 UP and DOWN regulated as my gene sets here. I made this list above.

index2<-ids2indices(gene.sets = CELLSV186DE, identifiers = identifiers)


#define a matrix of the contrast that you want the results for
contr<-makeContrasts(TreatV186.24-TreatMock.24, levels = design)

#put arguments into camera
V186cameraResult<-camera(y = y, index = index2, design = design, contrast = contr )

pander(V186cameraResult)
```
This is also weird and makes me think there is a mistake somewhere.


## Test 3
Now the Hallmark gene set from MSigDB (prepared for R here: http://bioinf.wehi.edu.au/software/MSigDB/, accessed 22Apr16). Note that ENTREZ_GENE_ID is the ID provided for these gene sets, so I'll have to use that for my identifiers.

More info on the MSigDB gene sets is here: http://software.broadinstitute.org/gsea/msigdb/collections.jsp#H

```{r,echo=FALSE,message=FALSE, warning=FALSE}

#load the Hallmark gene sets ("Hs.H")

load("MSigDB_gene_sets/human_Hallmark_v5.rdata")

##Expression set to use

y<-exprs(expressedProbes.lumi)


#IDs from eset: ENTREZ_IDS from the data I'm analyzing (not the pre-determined gene set). There are some duplicates in there but I am leaving them in for now because I know they will be removed non-randomly. This means less power though I think.

identifiers = as.character(fData(expressedProbes.lumi)[,"ENTREZ_GENE_ID"])

index3<-ids2indices(gene.sets = Hs.H, identifiers = identifiers)

#define a matrix of the contrast that you want the results for
contr<-makeContrasts(TreatV186.24-TreatMock.24, levels = design)

#put arguments into camera
hallmarkGSEA<-camera(y = y, index = index3, design = design, contrast = contr )

pander(head(hallmarkGSEA))

write.csv(hallmarkGSEA, file= "./CAMERA_gene_set_testing_output/hallmarkGSEA.csv")



```

## Test 4
Now trying with the immunologic gene sets ("c7") from MSigDB

```{r,echo=FALSE,message=FALSE, warning=FALSE}

#load the immunologic gene sets ("Hs.c7")

load("MSigDB_gene_sets/human_c7_v5.rdata")

##Expression set to use

y<-exprs(expressedProbes.lumi)


#IDs from eset: ENTREZ_IDS from the data I'm analyzing (not the pre-determined gene set). There are some duplicates in there but I am leaving them in for now because I know they will be removed non-randomly. This means less power though I think.

identifiers = as.character(fData(expressedProbes.lumi)[,"ENTREZ_GENE_ID"])

index4<-ids2indices(gene.sets = Hs.c7, identifiers = identifiers)

#define a matrix of the contrast that you want the results for
contr<-makeContrasts(TreatV186.24-TreatMock.24, levels = design)

#put arguments into camera
c7GSEA<-camera(y = y, index = index4, design = design, contrast = contr )

pander(head(c7GSEA))

write.csv(c7GSEA, file = "./CAMERA_gene_set_testing_output/c7GSEA.csv")


sessionInfo()

```