---
title: "Looking for housekeeping genes in human vaginal epithelial cell MA data"
output: github_document
---

We would like to find a good housekeeping gene to use for qPCR with HSV2 infected and un-infected human vaginal epithelial cells. Here I will analyze the results of the microarray experiment that Lamar did for the Herpes P01 project to look for probes with the following criteria:

* Probes for which there were at least 3 samples with detection p-values
that were <0.05. See documentation in my script that analyzes the HVE microarray data for rationale behind this choice.

* Probes with small difference in log2 fold change (<0.5) between mock and infected conditions.

* Probes with large p-value between mock and infected (p > 0.05)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(dplyr)
library(limma)
library(stringr)
library(reshape2)
library(ggplot2)
library(pander)
```


```{r V186.24   non DE}
#load data from the markdown_vaginal_cell_microarray.Rmd script. This data has already been filtered for detection. 
load("RdataOutput/CELLSfit2eBayes.Rdata")

# I am basically following the same steps here as I did for finding DE probes except I will filter for the opposite criteria, i.e. pvalue >0.05 and |lfc| <0.5.

#I only care about the 24hr timepoints for now.
#by not specifying a lfc and pval cutoff in the toptable call, I will get data for all probes (no filtering yet)
CELLSttV186.24all<-topTable(fit2,coef="V186.24vsMock.24", adjust.method = "BH", number=Inf)

# add a column indicating which contrast this is for

CELLSttV186.24all<-mutate(CELLSttV186.24all,contrast = "V186.24vsMock.24")

```


```{r SD90.24}

CELLSttSD90.24all<-topTable(fit2,coef="SD90.24vsMock.24", adjust.method = "BH", number=Inf)

# add a column indicating which contrast this is for

CELLSttSD90.24all<-mutate(CELLSttSD90.24all,contrast = "SD90.24vsMock.24")
```

```{r combine SD90 and V186 data}

CELLStt_186_SD90_all <- bind_rows(CELLSttSD90.24all,CELLSttV186.24all)


#now filter for just the probes with pval > 0.05 and |lfc|<0.5 (opposite of my criteria for DE probes)

CELLStt_186_SD90_nonDE<-CELLStt_186_SD90_all%>%
  filter(abs(logFC)<0.5,adj.P.Val>0.05 )


```


```{r only fisher endogenous controls}

#applied biosystems (fisher now I guess) offers 11 endogenous controls for qPCR:


# Here are the symbols and entrez IDs
# ACTB: 60
# RNA18s5:100008588
# GAPDH:2597
# PPIA:5478
# B2M: 567
# GUSB: 2990
# HPRT1: 3251
# PGK1: 5230
# RPLPO: 6175
# TBP: 6908
# TFRC: 7037 

#make a df of the symbols and IDS

endogCntls <-data.frame(symbol = c("ACTB", "RNA18s5","GAPDH","PPIA","B2M","GUSB","HPRT1","PGK1","RPLPO","TBP","TFRC"),
                        ENTREZ_GENE_ID = c("60", "100008588","2597","5478","567","2990","3251","5230","6175","6908","7037"))


#add a column to the nonDE top table saying whether or not the ENTREZ id is one of the endog cntls

CELLStt_186_SD90_nonDE<-CELLStt_186_SD90_nonDE %>%
                                   mutate(endog.cntl = (ENTREZ_GENE_ID %in% endogCntls$ENTREZ_GENE_ID))%>%
  arrange(endog.cntl)

# the gene for RNA18s5 shows up as a "LOC" gene instead of giving me the symbol I expect, but when I look up the LOC symbol in NCBI it takes me to human RNA18s5 so I guess that is right...
```

The following plot shows all probes that meet the adjusted p-value and log2 fold change criteria with the endogenous controls highlighted in red.

```{r plot probes}
#plot
ggplot(CELLStt_186_SD90_nonDE, aes(x = logFC,y = -log10(adj.P.Val)))+
  geom_point(aes(color = endog.cntl),alpha = .6)+
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))+
  facet_wrap(~contrast)+
  guides(color = FALSE)+
  ggtitle("V186 and SD90 24hrs: all probes with pvalue >0.05 and |lfc|<0.5")
```


## Log2 fold change and Adjusted P-values for common endogenous controls that met filtering criteria

```{r summary table}

cntlSummary<-CELLStt_186_SD90_nonDE %>%
  filter(endog.cntl == "TRUE")%>%
  select(contrast,TargetID,logFC, adj.P.Val)%>%
  arrange(TargetID)

#replace LOC100008588 with RNA18s5
cntlSummary$TargetID <-str_replace(cntlSummary$TargetID, "LOC100008588","RNA18s5")


pander(cntlSummary)


##Do the same procedure as for making a heat map to get exprs data for a certain contrast, then melt to make box plots. See SH's code
```

