
---
output:
  md_document:
    variant: markdown_github
---

"Vaginal cell herpes infection experiment"


```{r,echo=FALSE,message=FALSE, warning=FALSE}
require(dplyr)
require(lumi)
require(limma)

#READ IN RAW MICROARRAY DATA FINAL REPORT #######
#data sent from LMF

RAW<-"J:\\MacLabUsers\\HLADIK SHARED\\Projects\\Herpes study\\Herpes-Project-1\\vaginal_cell_microarray\\Finalreport_HSV2_HVE MA.txt"

RAW.lumi<-lumiR(RAW,detectionTh = 0.05, na.rm = TRUE,convertNuID = FALSE, dec = '.',
                parseColumnName = FALSE, checkDupId = FALSE,
                QC = TRUE,
                columnNameGrepPattern = list(exprs='AVG_SIGNAL',
                                                  se.exprs='BEAD_STDERR',
                                                       detection='Detection Pval',
                                                       beadNum='Avg_NBEADS'),
                inputAnnotation=TRUE,
                annotationColumn=c('ILMN_GENE', 'ENTREZ_GENE_ID', 'GI', 'ACCESSION', 'SYMBOL', 'PROBE_ID', 'PROBE_START', 'PROBE_SEQUENCE', 'CHROMOSOME', 'PROBE_CHR_ORIENTATION', 'PROBE_COORDINATES'),
                verbose = TRUE)
```

SOME PLOTS OF NON NORMALIZED DATA:
LMF left sample 3E out of the analysis, but it is still included here.
 Density plot, cdf plot, sample relations, boxplot. Looks like 3D and 6D are outliers, both are V186 24hrs.
 
```{r,echo=FALSE,message=FALSE, warning=FALSE}
#density plot
density(RAW.lumi)

#looking at JUST samples 3E, 6D, 3D
density(RAW.lumi[,25:27])

#3E doesn't look like an outlier here.

#CDF plot: cumulative probability of having <= a certain log2 intensity
plotCDF(RAW.lumi)

#sample relations
plot(RAW.lumi, what='sampleRelation',method="mds")

#boxplot
boxplot(RAW.lumi)


#adding in more complete pData that matches my formatting for the #explant data. 

pData <- read.table(".//vaginal_cell_microarray//vaginalCellMicroarrayPhenoData.txt",
                       sep="\t",row.names=1, header=TRUE)



#create metadata df
metadata<- data.frame(labelDescription = c("PTID","Sampling Time", "Treatment:either a virus strain or Mock."),
                      row.names=c("TissueID","Time","Treatment"))


#combine metadata and pdata into an annotated df
adf<-new("AnnotatedDataFrame",data=pData,varMetadata=metadata)

#create experiment data
experimentData<-new("MIAME",name="Claire Levy",
                    lab="Florian Hladik Lab",title="Vaginal Cell Microarray")


#make a lumiBatch that contains both the raw data we got from
#shared resources AND the phenoData, metadata and experiment data.
complete.RAW.lumi <- new("LumiBatch", exprs = exprs(RAW.lumi),phenoData = adf,
                  experimentData = experimentData,
                  se.exprs = se.exprs(RAW.lumi),
                  detection = detection(RAW.lumi))

save(complete.RAW.lumi,file = "CELLScomplete.RAW.lumi.Rdata")


###################BACKGROUND CORRECTION ####################
#the data we got from the core had no background correction when we did the explant experiment so I assume this wasn't corrected either.
B.complete.RAW.lumi<-lumiB(complete.RAW.lumi)


#################### VST TRANSFORMATION ######################
#"Stabilizing the expression variance based on
#the bead level expression variance and mean relations"

TB.complete.RAW.lumi <-lumiT (B.complete.RAW.lumi)

################## ROBUST SPLINE NORMALIZATION ################

NTB.complete.RAW.lumi<-lumiN(TB.complete.RAW.lumi,method="rsn")

################# QUALITY CONTROL ############################

QNTB.complete.RAW.lumi <- lumiQ(NTB.complete.RAW.lumi,detectionTh=0.05)

save(QNTB.complete.RAW.lumi, file = "CELLSQNTB.complete.RAW.lumi.Rdata")
```

PLOTS OF NORMALIZED DATA: all data, then just 3E, 6D and 3D


```{r,echo=FALSE,message=FALSE, warning=FALSE}

density(QNTB.complete.RAW.lumi)

density(QNTB.complete.RAW.lumi[,25:27])

plot(QNTB.complete.RAW.lumi, what='sampleRelation',method="mds")

plot(QNTB.complete.RAW.lumi[,25:27], what='sampleRelation',method="mds")

boxplot(QNTB.complete.RAW.lumi)

boxplot(QNTB.complete.RAW.lumi[,25:27])
```
FILTERING PROBES BASED ON DETECTION 
 Limma suggests to keep probes that are expressed above bg on 
at least n arrays where n is smallest number of replicates assigned
to any of the treatment combinations.

Our treatment combinations are TissueID/Treatment/Time
We have 3 Tissue IDs x 3 Treatments x 3 time points = 27 so the smallest number
 of replicates possible for any of those 3 arrays is 3

```{r,echo=FALSE,message=FALSE, warning=FALSE}
detectedProbes <- rowSums(detection(QNTB.complete.RAW.lumi)<0.05)>=3

#now extract just those probes that are TRUE from the lumibatch
#meaning there were at least 3 samples with detection p-values
#that were <0.05

expressedProbes.lumi <-QNTB.complete.RAW.lumi[detectedProbes,]
```

how many probes did we have before and after filtering?
```{r,echo=FALSE,message=FALSE, warning=FALSE}
dims(QNTB.complete.RAW.lumi)#47323

dims(expressedProbes.lumi)#22894
```
how many removed?
```{r,echo=FALSE,message=FALSE, warning=FALSE}
dims(QNTB.complete.RAW.lumi)-dims(expressedProbes.lumi)#20192

save(expressedProbes.lumi, file="CELLSexpressedProbes.lumi.Rdata")
```
Here's the design matrix for the analysis

```{r,echo=FALSE,message=FALSE, warning=FALSE}
############### TARGETS AND DESIGN MATRIX ###################
# see section 9.4.1 and 9.4.2 in the limma users guide

targets<-pData(expressedProbes.lumi)%>%
  select(TissueID, Treatment,Time)
TissueID<- factor(targets$TissueID)
Treat <-factor(paste(targets$Treatment,targets$Time, sep="."))

design<-model.matrix (~0+Treat+TissueID)
save(design, file="CELLSdesign.Rdata")
design
```
Here's the contrasts matrix

```{r,echo=FALSE,message=FALSE, warning=FALSE}
####################### FIT MODEL TO PROBES ###################
fit <- lmFit(expressedProbes.lumi,design=design)
save(fit, file="CELLSfit.Rdata")

#Now we can make any comparisons
#between the experimental conditions

# If topTable is called and coef has two or more elements,
# then the specified columns will be extracted from fit and
# topTableF called on the result. topTable with coef=NULL is 
# the same as topTableF, unless the fitted model fit has only
# one column.

bothcm<-makeContrasts(
  V186.3vsMock.3 = TreatV186.3-TreatMock.3,
  V186.8vsMock.8 = TreatV186.3-TreatMock.8,
  V186.24vsMock.24 = TreatV186.24-TreatMock.24,
  SD90.3vsMock.3 = TreatSD90.3-TreatMock.3,
  SD90.8vsMock.8 = TreatSD90.8-TreatMock.8,
  SD90.24vsMock.24 = TreatSD90.24-TreatMock.24,
  levels=design
)

bothcm

#fit the contrasts  
fit2<-contrasts.fit(bothcm, fit=fit)


#compute diff exprsn
fit2 <-eBayes(fit2)

save(fit2, file= "CELLSfit2eBayes.Rdata")
```

How many probes are up and down regulated for each contrast?

```{r,echo=FALSE,message=FALSE, warning=FALSE}

#method=separate is same as doing topTable for all coefs separately
results<-decideTests(fit2,method="separate", adjust.method="BH",
                      p.value=0.05, lfc=0.5)

save(results, file= "CELLSdecideTestsResults.Rdata")

#turn the results matrix into a data frame and make the
#probeID a real column and remove the rownames

resultsDF<-as.data.frame(results)
resultsDF$ProbeID<-rownames(resultsDF)
rownames(resultsDF)<-NULL

#melt the df for easy summarizing
library(reshape2)

resultsDFmelt<-melt(resultsDF, id.vars="ProbeID")
save(resultsDFmelt,file = "resultsDFmelt.Rdata")
#number of up and down regulated probes based on 
#p.val at least 0.05 and lfc at least 0.5

summary<-resultsDFmelt %>%
  group_by(variable)%>%
 summarize(down=sum(value=="-1"),up=sum(value=="1"))
summary
save(summary, file = "CELLSsummary.Rdata")
