

"Vaginal explant herpes infection experiment"

```{r,echo=FALSE,message=FALSE, warning=FALSE}
require(dplyr)
require(lumi)
require(limma)

#READ IN RAW MICROARRAY DATA FINAL REPORT #######
#data from shared resources

RAW<-"J:/MacLabUsers/HLADIK SHARED/Projects/Herpes study/Herpes-Project-1/Illumina/2015_11_10/GenomeStudioProject/SeanHughes_HumanHT12v4_151112/2015.11.12smhughesFinalReport.txt"

RAW.lumi<-lumiR(RAW,detectionTh = 0.05, na.rm = TRUE,convertNuID = FALSE, dec = '.',
                parseColumnName = FALSE, checkDupId = FALSE,
                QC = TRUE,
                columnNameGrepPattern = list(exprs='AVG_SIGNAL',
                                                       se.exprs='BEAD_STDERR',
                                                       detection='Detection Pval',
                                                       beadNum='Avg_NBEADS'),
                inputAnnotation=TRUE,
                annotationColumn=c('ILMN_GENE', 'ENTREZ_GENE_ID', 'GI', 'ACCESSION', 'SYMBOL', 'PROBE_ID', 'PROBE_START', 'PROBE_SEQUENCE', 'CHROMOSOME', 'PROBE_CHR_ORIENTATION', 'PROBE_COORDINATES'),
                verbose = TRUE)

load("RAW.lumi.Rdata")
```

EXPERIMENT NOTES 
6 samples failed when the microarray was run 6,33,34,35,45,63
and two failed QC (36,64). The failed were left out of the finalReport filethat we got from shared resources and we will leave all 8 out of the analysis too.

ALSO

The first 5 samples in the micrarray data are T cells that 
were exposed to Tenofovir. These will be analyzed separately and
can be separated from the vaginal explant data.


```{r,echo=FALSE,message=FALSE, warning=FALSE}
########### SUBSETTING THE LUMIBATCH ######################

#subset the lumibatch to remove the non-explant data AND the two samples
#that were left in the finalReport but failed the QC
RAW.lumi<-RAW.lumi[,-c(1:5,32,58)]#32 and 58 are the indices for
#samples "36" and "64"

#read in phenoData (does NOT include any phenoData for non-explant samples)

pData<- read.table("explantMicroarrayPhenoData.txt",
                       sep="\t",row.names=1, header=TRUE)




#make the rownames character(they are numbers, but are actually
#sample names and lumi has them as character)

rownames(pData)<-as.character(rownames(pData))
pData$TissueID<-as.character(pData$TissueID)
pData$Time <- as.character(pData$Time)


#need to remove phenodata for the samples that failed. The rownames
#start at 7 (1-6 were the tenofovir samples). So the sampleNames are now
#6 ahead of the the row numbers (sample 33 is in the 27th row of data)
# missing samples 33=ID326 T1 M D1
#34=ID 326 T2 V1 D1
#35=ID 326 T3 V2 D1
#36 ID 317 T2 M D1
#45 ID 318 T3 V1 D1
#63 ID 327 T2 V1 D1
#64 ID 319 T2 M D1

#make a vector of rownames I want to remove
toExclude<-c("33","34","35","36","45","63","64")

#to get the indices of the rownames that I want to remove:
#>which(rownames(pData)%in% toExclude)

#and then remove those indices from pData

pData<-pData[-which(rownames(pData)%in% toExclude),]



#create metadata df
metadata<- data.frame(labelDescription = c("Tissue,Time,Virus,Dose. Dose is same for all.",
                                           "PTID","Sampling Time","Virus: V1=strain SD90, V2=strain 186, M=Mock", "Treatment:either a virus strain or Mock.",
                                           "Location of original RNA on 96well plate"),
                      row.names=c("DescriptiveSampleID","TissueID","Time","Dose","Treatment","PlateID"))


#combine metadata and pdata into an annotated df
adf<-new("AnnotatedDataFrame",data=pData,varMetadata=metadata)

#create experiment data
experimentData<-new("MIAME",name="Claire Levy",
                    lab="Florian Hladik Lab",title="Vaginal Explant Microarray")


#make a lumiBatch that contains both the raw data we got from
#shared resources AND the phenoData, metadata and experiment data
#that I created MINUS the tenofovir data and the failed samples.

complete.RAW.lumi<-new("LumiBatch", exprs=exprs(RAW.lumi),phenoData=adf,
                  experimentData=experimentData,
                  se.exprs=se.exprs(RAW.lumi),
                  detection=detection(RAW.lumi))

save(complete.RAW.lumi,file="complete.RAW.lumi.Rdata")
```
 SOME PLOTS OF NON NORMALIZED DATA:
 density plot, cdf plot
```{r,echo=FALSE,message=FALSE, warning=FALSE}
#density plot
density(complete.RAW.lumi)#number of probes for each sample that occur
#at a certain log2 intensity

#CDF plot: cumulative probability of having <= a certain log2 intensity
plotCDF(complete.RAW.lumi)

#sample relations
plot(complete.RAW.lumi, what='sampleRelation',method="mds")

#boxplot
boxplot(complete.RAW.lumi)

###################BACKGROUND CORRECTION ####################
#the data we got from the core had no background correction so I will do it here

B.complete.RAW.lumi<-lumiB(complete.RAW.lumi)


#################### VST TRANSFORMATION ######################
#"Stabilizing the expression variance based on
#the bead level expression variance and mean relations"

TB.complete.RAW.lumi <-lumiT (B.complete.RAW.lumi)

################## ROBUST SPLINE NORMALIZATION ################

NTB.complete.RAW.lumi<-lumiN(TB.complete.RAW.lumi,method="rsn")

################# QUALITY CONTROL ############################

QNTB.complete.RAW.lumi <- lumiQ(NTB.complete.RAW.lumi,detectionTh=0.05)

save(QNTB.complete.RAW.lumi, file = "QNTB.complete.RAW.lumi.Rdata")
```
PLOTS OF NORMALIZED DATA 

```{r,echo=FALSE,message=FALSE, warning=FALSE}
plot(QNTB.complete.RAW.lumi)

plot(QNTB.complete.RAW.lumi, what='sampleRelation',method="mds")

boxplot(QNTB.complete.RAW.lumi)

```
FILTERING PROBES BASED ON DETECTION 
 Limma suggests to keep probes that are expressed above bg on 
at least n arrays where n is smallest number of replicates assigned
to any of the treatment combinations.

Our treatment combinations are TissueID/Treatment/Time
We have 7 Tissue IDs x 3 Treatments x 3 time points=63 so the smallest number
 of replicates possible for any of those 63 arrays is 7

```{r,echo=FALSE,message=FALSE, warning=FALSE}
#this is how the detection filtering works
#Mark the detection pvalues (there is one per probe per sample) with a 1 if <0.05 or a 0 if >0.05
# using (detection(QNTB.complete.RAW.lumi)<0.05)

#add up the 0's and 1's across each row (i.e. for all the samples)
#using rowSums

#now you have the number of detection p values <0.05 for each probe
#for all the samples (max possible = total samples)

#now tell me which probes have a rowSum of >=7 (probes will be marked as TRUE
# or false if they do or do not have >=7 rowSum)
            
detectedProbes <- rowSums(detection(QNTB.complete.RAW.lumi)<0.05)>=7

#now extract just those probes that are TRUE from the lumibatch

expressedProbes.lumi <-QNTB.complete.RAW.lumi[detectedProbes,]
```

how many probes did we have before and after filtering?
```{r,echo=FALSE,message=FALSE, warning=FALSE}
dims(QNTB.complete.RAW.lumi)#47323

dims(expressedProbes.lumi)#21726
```
how many removed?
```{r,echo=FALSE,message=FALSE, warning=FALSE}
dims(QNTB.complete.RAW.lumi)-dims(expressedProbes.lumi)#20192

save(expressedProbes.lumi, file="expressedProbes.lumi.Rdata")

############### TARGETS AND DESIGN MATRIX ###################
# see section 9.4.1 and 9.4.2 in the limma users guide

targets<-pData(expressedProbes.lumi)%>%
  select(TissueID, Treatment,Time)
TissueID<- factor(targets$TissueID)
Treat <-factor(paste(targets$Treatment,targets$Time, sep="."))

design<-model.matrix (~0+Treat+TissueID)

####################### FIT MODEL TO PROBES ###################
fit <- lmFit(expressedProbes.lumi,design=design)


#Now we can make any comparisons
#between the experimental conditions

# If topTable is called and coef has two or more elements,
# then the specified columns will be extracted from fit and
# topTableF called on the result. topTable with coef=NULL is 
# the same as topTableF, unless the fitted model fit has only
# one column.

bothcm<-makeContrasts(
  V186.3vsMock.3 = TreatV186.3-TreatMock.3,
  V186.8vsMock.8 = TreatV186.3-TreatMock.8,
  V186.24vsMock.24 = TreatV186.24-TreatMock.24,
  SD90.3vsMock.3 = TreatSD90.3-TreatMock.3,
  SD90.8vsMock.8 = TreatSD90.8-TreatMock.8,
  SD90.24vsMock.24 = TreatSD90.24-TreatMock.24,
  levels=design
)


#fit the contrasts  
fit2<-contrasts.fit(bothcm, fit=fit)


#compute diff exprsn
fit2 <-eBayes(fit2)


```

How many probes are up and down regulated for each contrast?
```{r,echo=FALSE,message=FALSE, warning=FALSE}

#method=separate is same as doing topTable for all coefs separately
results<-decideTests(fit2,method="separate", adjust.method="BH",
                      p.value=0.05, lfc=0.5)

#turn the results matrix into a data frame and make the
#probeID a real column and remove the rownames

resultsDF<-as.data.frame(results)
resultsDF$ProbeID<-rownames(resultsDF)
rownames(resultsDF)<-NULL

#melt the df for easy summarizing
library(reshape2)
resultsDFmelt<-melt(resultsDF, id.vars="ProbeID")

#number of up and down regulated probes based on 
#p.val at least 0.05 and lfc at least 0.5

summary<-resultsDFmelt %>%
  group_by(variable)%>%
 summarize(down=sum(value=="-1"),up=sum(value=="1"))
summary


## topTable and heat maps of probes with 
#adj p values <0.05 and lfc >=0.5

tt<-topTable(fit2,coef=1, adjust.method = "BH", number=Inf, p.value=0.05, lfc=0.5)

selected<-rownames(tt)

esetSel<-expressedProbes.lumi[selected,]

heatmap(exprs(esetSel))

#looking and # of sig probes with different p val limits
x<- log10(tt$adj.P.Val)

plot(x, type="l")
sigline <- c(.05, .01, .005, .001,.0005, .0001)
sigline <- log10(sigline)
sigcolors <- c("red", "blue", "green", "yellow","pink","purple")
sapply(1:length(sigline), function(x){abline(h=sigline[x], col=sigcolors[x])})

```
Looks like there would still be a lot (hundreds)of DE probes for the v186 24hr comparison if reduced the adj p val cutoff to 0.01, but that would reduce the # of DE probes in the v186.3 to ~2 and v186.8 to ~3. The SD90.3 would be 2 and SD90.8 and .24 were already zero with p<=0.05

```{r,echo=FALSE,message=FALSE, warning=FALSE}
#getting some go terms

#add entrez ids to the melted p val and lfc filtered df
ProbeIDandSymbol<-fData(RAW.lumi)[,c(1,4,7)]

interestingProbes<-merge(resultsDFmelt,ProbeIDandSymbol, by="ProbeID")
#remove the unchanged probes
interestingProbes<-filter(interestingProbes, value!="0")

save(interestingProbes,file="interestingProbes.Rdata")

contrastList<-list(levels(resultsDFmelt$variable))

library(stringr)
get
str_detect_all(resultsDFmelt$variable,contrast




library(illuminaHumanv4.db)
library(biomaRt)
